Polymethylation scores paper

### 2021-10-11 Papers/Polymethylation_scores/04_background.Rmd
work on intro and background outline

# Background

P1: Maternal prenatal smoking is a public health concern but measuring prenatal maternal smoking behavior is challenging. Developing an in vitro biomarker for prenatal smoke exposure would be beneficial for personalized medicine.

P2: DNA methylation could provide a reliable and specific biomarker of prenatal maternal smoke exposure. Cordblood and peripheral blood DNA methylation has been significantly associated with prenatal maternal smoking. **Gap**: Most previous studies have used blood samples, but saliva is an easier to collect tissue. To our knowledge, no study has measured associations between prenatal maternal smoking and DNA methylation in saliva samples.

P3: DNA methylation has been associated with prenatal maternal smoking at various cross-sectional time points across the life course. Ideally, a biomarker of a prenatal exposure would persist across the life course. **Gap**: Few studies have compared the effectiveness of DNA methylation as biomarker of prenatal maternal smoking in the same individuals at different times.

P4: DNA methylation has been well replicated in primarily European ancestry cohorts. The performance of other biomarkers and risk scores varies across ancestry groups. To be applied clinically without exacerbating health disparities, it is important to understand how the perfomance of biomarkers vary across ancestry groups. **Gap**: Yet few (no?) studies have compared associations of prenatal maternal smoking across diffferent ancestry groups.

P5: Choice of how to construct a DNA methylation based biomarker.

P6: To translate DNA methylation to a *in vitro* biomarker for prenatal exposures, we must understand how performance [check with kelly if ok to use this word here now that have ROC plots] varies across tissue, age and ancestry. **To address these gaps we used XXXX longitudinal saliva samples from 9 and 14 year old children enrolled in a  diverse US birth cohort to test associations between DNA methylation summary measures and prenatal maternal smoke exposure.**  We found that maternal prenatal smoking was significantly associated with polymethylation scores. This association was persistent across time-point and consistent across ancestry strata. Polymethylation scores outperformed other non-target methylation summary measures, including global methylation and methylation clocks, as well as individual, *a priori* selected CpG sites. Our results suggest that polymethylation scores can serve as a persistent and consistent biomarker of prenatal smoke exposure.

>>>>

---
output:
  word_document: default
  html_document: default
---
# Background

Maternal prenatal smoking is a public health concern: in 2018, 11% of pregnant US women aged 15-44 reported any past-month cigarette use, and exposed children are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma. Measuring prenatal maternal smoking behavior is challenging.  Maternal prenatal smoking is underreported due to stigma. The gold-standard for smoking measurements, serum cotinine, has a short half-life of nine hours in pregnant women.   Further, serum cotinine measures during pregnancy are rarely available outside of birth cohorts, limiting clinical application as a marker of prenatal exposures. Developing a persistent, reliable and feasible biomarker of prenatal maternal smoke exposure is critical for advancing research and precision health.

DNA methylation has been associated with prenatal maternal smoking at various cross-sectional time points across the life course in primarily European ancestry groups. Cordblood, peripheral blood and placental DNA methylation has been significantly associated with prenatal maternal smoking. [Joubert 2016; Bustamante 2021; Davey Smith 2018]. Blood DNA methylation signatures in cohorts of older children and adults also were significantly associated with prenatal maternal smoking exposure [Terry 2019, Davey Smith 2018]. Yet few studies have examined associations at different ages in the same participants. Additionally, while associations are well-replicated in cohorts of European ancestry, few studies have used non-European ancestry populations and those that have had small sample sizes: Fallin 2016 186 individuals of African or admixed ancestry, Terry 2018 60% African or hispanic but only 89 participants, Burchard 2019 572 Latino children]. DNA methylation is associated with prenatal maternal smoking, but the consistency of this association across age and ancestry is understudied.

The association between prenatal maternal smoking and DNA methylation suggests DNA methylation could provide a reliable and specific biomarker of prenatal maternal smoke exposure. Importantly, changes in DNA methylation are specific to smoke exposure *in utero*. Previous work has found reasonable performance of cord blood based [London 2017] (accuracy 91%, sens 58%, spec 97%) [Fallin 2016] (81% accuracy) and peripheral blood based biomarkers (AUC: 0.72) [Davey Smith 2018] of prenatal maternal smoking.A 2018 study found that DNA methylation based scores could even predict predict prenatal smoke exposure (30 years previously) with an area under the curve of 0.72 (95% confidence interval 0.69, 0.76) [Davey Smith 2018]. However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across age and ancestry has not been evalutated. Few studies have compared the effectiveness of DNA methylation as biomarker of prenatal maternal smoking in the same individuals at different times or evaluated if the inclusion of multiple time points improves performance. Additionally, previous work has relied on blood and placental samples. Cord blood or placenta samples are unlikely to be available in a clinical setting. While peripheral blood is a more realistic clinical sample,  saliva is even easier to collect. To our knowledge, no study has evaluated salivary DNA methylation as a biomarker for prenatal maternal smoking.

DNA methylation changes hold promise as a biomarker for prenatal maternal smoking. However, before translation to clinical applications, we must understand how and if the signal varies across age, ancestry and tissue sample. The goal of this study was to assess the *in vitro* potential of salivary DNA methylation biomarkers for prenatal maternal smoking.  We tested associations between prenatal maternal smoke exposure and saliva DNA methylation biomarkers from the Illumina 450K chip, including epigenetic clocks, polymethylation scores, and individual *a priori* CpG sites using longitudinal samples from the Fragile Families and Child Wellbeing study. We performed age- and ancestry-stratified analyses to test the hypothesis that DNA methylation could serve as a consistent biomarker of prenatal maternal smoking across populations and time.

### 2021-10-11 Papers/Polymethylation_scores/06_discussion.Rmd
work on intro and background outline

Restate findings

How are results similar or different to other results

Tissue

Age

Ancestry

>>>>

P1 (restate findings): In this longitudinal birth cohort, we found an association between prenatal maternal smoking and salivary DNA methylation.

- Polymethylation scores were associated with prenatal maternal smoking across ages and ancestry while global methylation and epigenetic clocks were not.

- Polymethylation scores had reasonable accuracy and accuracy was more consistent across time when using polymethylation scores than when using individual *a priori* CpG sites.

P2 (compares to previous findings): Our findings fit within the previous literature on DNA methylation as a possible biomarker of prenatal maternal smoking.

- {if including EWAS results: many top hits from previous EWAS analyses using blood samples were replicated} (joubert 2016).

- Fits well with previous findings of an association in cord and peripheral blood (Joubert 2016). Although no studies of prenatal maternal smoking and saliva methylation, own-smoking and DNA methylation association in saliva, including at cg05575921 (Philibert 2019, Long 2019).

- Performance was worse than the performance of biomarkers using cord blood (AUC: 0.82-0.97 from Fallin 2016 London 2017), but comparable to performance of biomarkers using peripheral blood (AUC 0.72, Davey Smith 2018... NOTE: I don't think this was adjusted for anything else, i.e.  other predictors were included in the model--so technically then ours is worse).

- For own smoking saliva DNA methylation at cg05575921 has been found to have a higher accuracy (in models incorporating age gender and methylations status) was 0.995 in blood, 0.971 in saliva, improved to 0.981 when incorporating another methylation marker of tissue heterogeneity [Philibert, Long 2019]

P3 (what we add): We contribute to the translation of DNA methylation to a clinical biomarker by evaluating consistency across ages and ancestries.

- Even when polymethylation score coefficients were based on coefficients from a metaanalysis of primarily European ancestry, we found an association with prenatal maternal smoking. Effect estimates relatively consistent across ancestries

- AUC for polymethylation scores calculated using coefficients from models adjusting for cell-type better than those not adjusting for cell-type, cell-type control should be instituted both when creating weights and when applying scores to new population.

- AUC better for polymethylationn scores than cg055, and more consistent across ages. Better to incorporate multiple methylation sites than single methylation site?

P4 (limitations & strengths):

- Available coefficients for construction of scores were from blood, primarily European ancestry meta-analysis

- Exposure measurement not ideal: Used maternal self report as opposed to serum cotinine levels as exposure measurement - exposure misclassification/information bias; any vs sustained smoking

- Possibility for selection bias into cohort due to live birth bias

- Residual confounding is always possible.

- Large sample size and longitudinal measures

### 2021-10-11 Papers/Polymethylation_scores/index_paper.Rmd
worked on outlining introduction
work on intro and background outline

---
title: "Polymethylation scores paper"
author: "F blostein"
date: "9/13/2021"
output: pdf_document
bib: Methylation_SmokingBib.bib
---

>>>>

---
title: "Polymethylation scores paper"
author: "F blostein"
date: "9/13/2021"
output:
  word_document: default
  html_document:
    df_print: paged
bib: ../Methylation_SmokingBib.bib
---
